Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-01-29 parallel-tractability-J-v9/background.tex
splitting the main file for better organization

%\documentclass[preprint,12pt]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
 \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
\usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm}

\usepackage{amsmath}

\usepackage{CJK}
\usepackage{color}

\usepackage{enumitem}
\usepackage{rotating}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

%\journal{Nuclear Physics B}

\hyphenation{data-log}

\begin{document}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{Parallel Tractability of Ontology\\
Materialization: Technique and Practice}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \address[label1]{}
%% \address[label2]{}

\author[add1]{Zhangquan Zhou \corref{cor1}}
\ead{quanzz1129@gmail.com}
\cortext[cor1]{Corresponding author}

\author[add1]{Guilin Qi}
\ead{gqi@seu.edu.cn}

\author[add2]{Birte Glimm}
\ead{birte.glimm@uni-ulm.de}

\address[add1]{
    School of Computer Science, Southeast University, China
}

\address[add2]{
    Institute of Artificial Intelligence, University of Ulm
}

\begin{abstract}
  Materialization is an important reasoning service for many
  ontology-based applications, but the rapid growth of semantic data
  poses the challenge to efficiently perform materialization on
  large-scale ontologies. Parallel materialization algorithms work
  well for some ontologies, which is not always explained by the
  theoretical results. For example, for the YAGO ontology
  parallelization works well, although the reasoning problem for the
  used ontology language is not in \texttt{NC}, i.e.\ the theoretical
  complexity class for \emph{parallel tractability}. This motivates us
  to study the problem of \emph{parallel tractability} of ontology
  materialization from a theoretical perspective. We focus on the
  datalog rewritable ontology languages DL-Lite and Description Horn
  Logic (DHL) and propose algorithms, called \texttt{NC} algorithms,
  to identify classes of ontologies for which materialization is
  tractable in parallel. To verify the practical usability of the
  above results, we analyze real-world datasets and show that for many
  ontologies expressed in DHL materialization is tractable in
  parallel. The implementation of our optimized parallel algorithm
  shows performance improvements over the highly optimized
  state-of-the-art reasoner RDFox on ontologies for which
  materialization is tractable in parallel.
\end{abstract}

\begin{keyword}
ontology \sep materialization \sep datalog \sep parallel tractability \sep NC complexity
\end{keyword}

\end{frontmatter}

%% \linenumbers

\section{Introduction}
\label{sec:introduction}

The Web Ontology Language (OWL)\footnote{The latest version is OWL 2: http://www.w3.org/TR/owl2-overview/}
is an important standard for ontology language in the Semantic Web and knowledge-based applications.
In these applications, \emph{materialization}, which is the reasoning task of computing all implicit
facts that follow from a given ontology \cite{handbook}, plays an important role. By means of it, developers and users
can optimize query answering, ontology diagnosis and debugging. Since huge amounts of semantic data
are being generated at an increasing pace by sensor networks, government authorities and social
media \cite{LehmbergRMB16,MeuselBP15}, it is challenging to conduct materialization on such large-scale ontologies efficiently.

There has been work proposing parallel reasoning algorithms and employing parallel computing platforms
for the ontology language RDFS and its extended
fragments \cite{MotikNPHO14,PetersSZ15,SubercazeGCL16}. Several optimization strategies, e.g., dictionary encoding, balancing workload
and data partitioning, are further studied to enhance parallel RDFS reasoning. There are also parallel implementations for scalable
reasoning of highly expressive ontology languages \cite{SteigmillerLG14,WuH12}. On the other hand, several work utilizes different kinds of computing platforms to make reasoning tasks more efficient in parallel, such as supercomputers \cite{Hoeksema2011,GoodmanJMAAH11},
MapReduce \cite{UrbaniKMHB12} and GPU servers \cite{HeinoP12}.

The above work verifies the effectiveness of parallel reasoning based on the evaluations on different test datasets.
However, for most popular ontology languages, even RDFS and datalog rewritable ontology languages,
which have \texttt{PTime}-complete or higher complexity of reasoning in the worst case,
they are not parallelly tractable according to \cite{Raymond95}, i.e., the efficiency of reasoning may not be
improved on a parallel implementation. A possible reason of the high performance of parallel
reasoning is that the utilized test datasets do not fall in the worst cases in terms of computational complexity.
It has also been shown that some well-known large-scale ontologies, such as YAGO, have good performance for parallel
reasoning \cite{KolovskiWE10}, but they are expressed in ontology languages that are not parallelly tractable in theory.
On the other hand, there do exist ontologies for which materialization cannot
always be improved by parallelism (we use experiments to show this in Section~\ref{sec:evaluation}).
The current work of parallel ontology reasoning can hardly explain this.
While one can try out different parallel implementations to see whether an ontology can be handled by (one of) them efficiently,
we study the problem of parallel tractability in theory and identify properties that make an ontology parallelly tractable.
These properties can be further used to optimize parallel algorithms and to guide ontology engineers in creating ontologies
for which parallel tractability can be guaranteed theoretically.

According to \cite{MotikNPHO14}, many real large-scale ontologies are essentially expressed in the ontology languages
that can be rewritten into datalog rules. Thus, we focus on such datalog rewritable ontology languages in this paper.
The main target of this paper is to identify the classes of datalog rewritable ontologies such that materialization
over these ontologies is parallelly tractable, i.e., in the parallel complexity class NC \cite{Raymond95}.
This complexity class consists of problems that can be solved efficiently in parallel.
To show that a problem is in the NC class, one can give an \emph{NC algorithm} that handles this problem
in parallel computation \cite{Raymond95}.
An NC algorithm is required to terminate in parallel poly-logarithmic time.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not NC algorithms,
since they are designed for general datalog programs and has \texttt{PTime}-complete complexity.
Thus, we first give NC algorithms that perform materialization,
and then identify the corresponding classes of datalog rewritable ontologies (called \emph{parallelly tractable classes})
that can be handled by these NC algorithms. To make the proposed NC algorithms practical,
we also discuss how to optimize and implement them.

In the practical part of this work, we study specific datalog rewritable ontology languages.
We first focus on the ontology language DL-Lite to clarify how to apply the above NC algorithms
to study the parallel tractability of ontology materialization.
We show that DL-Lite$_{core}$ and DL-Lite$_\mathcal{R}$ are parallelly tractable.
We then study the ontology language Description Horn Logic (DHL) \cite{GrosofHVD03}, which is the intersection of datalog and OWL
in terms of expressivity. We give a case of a DHL ontology where materialization can hardly be parallelized.
Based on the analysis of this case, we propose to restrict the usage of DHL such that materialization over the
restricted ontologies can be handled by the proposed NC algorithms.
We further extend the results to an extension of DHL that also allows complex role inclusion axioms.
Finally, we analyze well-known benchmarks and real-world datasets and show that many real ontologies following the
proposed restrictions belong to the parallelly tractable classes.
We implement a system based on an optimized NC algorithm for DHL materialization.
We compare our system to the state-of-the-art reasoner RDFox.
The experimental results show that the optimization proposed in this paper results in a better performance on parallelly
tractable ontologies compared with RDFox.

The rest of the paper is organized as follows. In Section~\ref{sec:background}, we introduce some basic notions.
We then give two NC algorithms for ontology materialization in Section~\ref{sec:ptclass}.
We study the parallelly tractable materialization of DL-Lite, DHL and an extension of DHL in Section~\ref{sec:ptonto} respectively.
In Section~\ref{sec:practicalAlg},
we discuss how to optimize and implement the given NC algorithms.
We analyze real-world datasets and evaluate our implementation in Section~\ref{sec:evaluation}.
We then discuss related work in Section~\ref{sec:related} and conclude in Section~\ref{sec:conclusion}.
The detailed proofs of the theorems and the lemmas in this paper are given in the appendix.

@@

\section{Parallel Tractability of Datalog Programs}
\label{sec:ptclass}

Our target is to find for which kinds of ontologies (not ontology languages)
materialization is parallelly tractable.
Since we assume that for any class of datalog programs where the rule
set of each datalog program is fixed, the materialization problem is thus in data complexity
$\texttt{PTime}$-complete, which is considered to be inherently sequential in the worst
case \cite{Raymond95}. In other words, the materialization problem on general
datalog programs cannot be solved in parallel poly-logarithmic time unless P=NC.
Thus, we say that \emph{materialization on a class of datalog programs is parallelly tractable
if there exists an algorithm that handles this class of datalog programs and runs in parallel
poly-logarithmic time (this algorithm is also called an NC algorithm)}.
In this section, we identify such classes of datalog programs by studying different NC algorithms.

\subsection{Parallel Tractable Classes}

We first give the following definition for a parallel tractable class of datalog programs
where an NC algorithm exists for handling each datalog program in this class.

\begin{definition}\label{def:ptd}
\textbf{(Parallelly Tractable Class)} Given a class $\mathcal{D}$ of datalog programs where the
same rule set is shared, we say that $\mathcal{D}$ is a parallelly tractable datalog program (\texttt{PTD}) class
if there exists an NC algorithm that performs sound and complete materialization for each datalog program
in $\mathcal{D}$. The corresponding class of ontologies of $\mathcal{D}$ is called a
parallelly tractable ontology (\texttt{PTO}) class.
\end{definition}

According to the above definition, if we find an NC algorithm $\mathsf{A}$
for datalog materialization, then we can identify a \texttt{PTD} class $\mathcal{D}_{\mathsf{A}}$,
which is the class of all datalog programs that can be handled by $\mathsf{A}$.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not NC algorithms
due to their \texttt{PTime}-complete complexity, since they are designed for handling general datalog programs.
Thus we give our NC algorithms.
In the following, we first give a parallel materialization algorithm that works for
general datalog programs. We then restrict this algorithm to an NC version
and identify the target \texttt{PTD} class.

\subsection{Materialization Graph}

In order to give a parallel materialization algorithm,
we introduce the notion of \emph{materialization graph}.
It makes the analysis of the given algorithm convenient.

\begin{definition}
\textbf{(Materialization Graph)}\label{def:mg}
A \emph{materialization graph}, with respect to
a datalog program $P=\langle R, \textbf{I}\rangle$, is a directed acyclic graph
denoted by $\mathcal{G}=\langle V, E\rangle$ where,
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $V$ is the node set and $V\subseteq T_R^{\omega}(\textbf{I})$;
  \item $E$ is the edge set and $E\subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$.
\end{enumerate}
For each edge of the form $e(v_1, v_2)$ where $v_1$ and $v_2$ are the two nodes,
we say that node $v_1$ is the parent (node) of node $v_2$ and node $v_2$ is the child (node)
of node $v_1$. Further, $\forall H,B_1,...,B_n\in V$, $\mathcal{G}$ satisfies the following conditions:
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $H$ has a parent node or a child node;
  \item $H$ is the original fact of $P$ if $H$ has the in-degree of $0$;
  \item $B_1,...,B_n\rightarrow H\in P^*$ is satisfied if $e(B_1, H),...,e(B_n, H)\in E$ and
  $B_1,...,B_n$ are \d{all} the parents of $H$.
\end{enumerate}
\end{definition}

For some derived atom $H$, there may exist several rule instantiations where $H$
occurs as a head atom.
This also means that $H$ can be derived in different ways.
The condition in the definition above results in only one way of deriving $H$ being
described by a materialization graph.
Suppose $\mathcal{G}$ is a materialization graph, the nodes whose in-degree is 0 are the
original facts in $\textbf{I}$.
The size of $\mathcal{G}$, denoted by $|\mathcal{G}|$, is the number of nodes in $\mathcal{G}$.
The depth of $\mathcal{G}$, denoted by \texttt{depth}($\mathcal{G}$), is the maximal length of a path
in $\mathcal{G}$.
We next give an example of a materialization graph.\\

\begin{example}\label{exp:mg}
Consider a DHL($\circ$) ontology $\mathcal{O}_{ex_1}$ where the TBox is
$\{\exists R.A\sqsubseteq A\}$, the RBox is $\{S\circ R\sqsubseteq R\}$ and
the ABox is $\{A(b),R(a_1,b),S(a_i,a_{i-1})\}$ for $2\leq i\leq k$ and
$k$ is an integer greater than $2$.
The corresponding datalog program of this ontology is $P_{ex_1}=\langle R, \textbf{I}\rangle$
where $\textbf{I}$ contains all the assertions in the ABox
and $R$ contains the two rules `$R(x,y),A(y)\rightarrow A(x)$'
and `$S(x,y),R(y,z)\rightarrow R(x,z)$'.
The graph in Figure~\ref{fig:mg} is a materialization graph with respect to $P_{ex_1}$,
denoted by $\mathcal{G}_{ex_1}$.
The nodes whose in-degree is 0 are the original facts in $\textbf{I}$;
each of the other nodes corresponds to a ground instantiation of some rule.
For example, the node $A(a_k)$ corresponds to the ground rule instantiation
`$R(a_k,b),A(b)\rightarrow A(a_k)$'.
The size of this materialization graph is the number of nodes, that is $3k$.
The depth of $\mathcal{G}_{ex_1}$ is $k$.
\end{example}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.7\textwidth]{fig-mg.eps}
\caption{An example of a materialization graph.}
\label{fig:mg}
\end{center}
\end{figure}

We say that a materialization graph $\mathcal{G}$ is a \emph{complete materialization graph}
when $\mathcal{G}$ contains all ground atoms in $T_R^{\omega}(\textbf{I})$.
The set of nodes in a complete materialization graph is actually the result
of materialization.
Thus, the procedure of materialization can be transformed to the construction of
a complete materialization graph.
We pay our attention to complete materialization graphs and do not distinguish it to the
notion `materialization graph'.
It should also be noted that there may exist several materialization graphs for a datalog program.

\subsection{A Basic Parallel Algorithm}
\label{sec:alg-bsc}

In this part, we propose a parallel algorithm
(denoted by Algorithm~$\mathsf{A}_{bsc}$) that constructs a materialization graph for a given datalog program.
Our computational model is a PRAM (parallel, random-access machine) that is mostly used to analyze the parallel
complexity. This model allows us giving \emph{the parallel assumption}:
for any datalog program $P=\langle R, \textbf{I}\rangle$,
any substitution of some atom and any rule instance in $P^*$ can be mapped to a unique memory
location; further, a one-to-one relation can be established between processors and rule instances.
Under this assumption, a processor can check the applicability of its corresponding
rule instance and access the state of an atom occurring in this rule instance in constant time.
Algorithm~$\mathsf{A}_{bsc}$ is then given as follows.\\

\noindent\texttt{Algorithm~$\mathsf{A}_{bsc}$}. Given a datalog program $P=\langle R, \textbf{I}\rangle$,
the algorithm returns a materialization graph $\mathcal{G}$ of $P$.
Suppose we have $|P^*|$ processors, and each rule instantiation in $P^*$ is
assigned to one processor.
Initially $\mathcal{G}$ is empty. The following three steps are then performed:
\begin{enumerate}[leftmargin=8ex,label=(\textit{Step \arabic*}),ref=Step~\arabic*]
\item Add all facts in $\textbf{I}$ to $\mathcal{G}$.\label{alg1:addFacts}
\item For each rule instantiation $B_1,...,B_n\rightarrow H$, if the body atoms are all
    in $\mathcal{G}$ while $H$ is not in $\mathcal{G}$,
    the corresponding processor adds $H$ to $\mathcal{G}$ and creates edges pointing
    from $B_1,...,B_n$ to $H$.\label{alg1:updateG}
\item If no processor can add more nodes and edges to $\mathcal{G}$, terminate, otherwise iterate \ref{alg1:updateG}.\label{alg1:halt}\hfill$\Box$
\end{enumerate}

\begin{example}
We consider the datalog program $P_{ex_1}$ in Example~\ref{exp:mg} again,
and perform Algorithm~$\mathsf{A}_{bsc}$ on it.
Initially, all the facts ($A(b),R(a_1,b),S(a_2,a_1),...,S(a_{k},a_{k-1})$) are added to the
result $\mathcal{G}_{ex_1}$ (\ref{alg1:addFacts}).
Then in different iterations of \ref{alg1:updateG}, the remaining nodes are added to
$\mathcal{G}_{ex_1}$ by different processors.
For example a processor $p$ is allocated a rule instantiation `$R(a_2,b),A(b)\rightarrow A(a_2)$'.
Then, processor $p$ adds $A(a_2)$ to $\mathcal{G}_{ex_1}$ after it checks that
$A(b)$ and $R(a_2,b)$ are in $\mathcal{G}_{ex_1}$.
Algorithm~$\mathsf{A}_{bsc}$ halts when $A(a_k)$ has been added to $\mathcal{G}_{ex_1}$ (\ref{alg1:halt}).
\end{example}

Lemma~\ref{lemma:a1} shows the correctness of Algorithm~$\mathsf{A}_{bsc}$ and that, for any datalog program $P$,
Algorithm~$\mathsf{A}_{bsc}$ always constructs a materialization graph with the minimum depth among all the
materialization graphs of $P$. The detailed proofs of Lemma~\ref{lemma:a1} and other lemmas and theorems can be found
in the appendix.

\begin{lemma}
\label{lemma:a1}
Given a datalog program $P=\langle R, \textbf{I}\rangle$, we have
\begin{enumerate}[leftmargin=4ex]
\item Algorithm~$\mathsf{A}_{bsc}$ halts and returns a materialization graph $\mathcal{G}$ of $P$;
\item $\mathcal{G}$ has the the minimum depth among all the materialization graphs of $P$.
\end{enumerate}
\end{lemma}

\begin{proof}[Proof sketch] This lemma can be proved by performing
an induction on $T_R^{\omega}(\textbf{I})$.
The stage (see the related contents in Section~\ref{sec:background}) of $P$
is the lower-bound of the depth of the materialization graphs. Based on the previous induction,
one can further check that, for the materialization graph $\mathcal{G}$ constructed by Algorithm~$\mathsf{A}_{bsc}$,
its depth equals the depth of the stage.
\end{proof}

We now discuss the parallel complexity of Algorithm~$\mathsf{A}_{bsc}$.
Given a class of datalog programs $\mathbb{P}$ where
a rule set is shared for each datalog program $P=\langle R, \textbf{I}\rangle$ in $\mathbb{P}$.
Let $e$, $v$ and $r$ represent the maximum arity of any EDB predicate, the maximum number of variables
in any datalog rule, and the number of datalog rules respectively.
We then have that the number of
constants is at most $|\textbf{I}|e$, and the number of all possible rule instances in $P^*$
is at most $r(|\textbf{I}|e)^v$.
Note that $e, v$ and $r$ depend only on the rule set $R$ and not on the fact set $\textbf{I}$.
Thus, the memory space for storing the atoms and the rule instances is polynomial in the size of $\textbf{I}$.
This also means that the number of processors is polynomially bounded.
The computing time of \ref{alg1:addFacts} and \ref{alg1:halt} occupy constant time (denoted by $c_1$) because of parallelism.
Since Algorithm~$\mathsf{A}_{bsc}$ works under the parallel assumption, one iteration of
\ref{alg1:updateG} also costs constant time (denoted by $c_2$). Thus,
the whole computing time of Algorithm~$\mathsf{A}_{bsc}$ turns out to be $c_1+l\cdot c_2$
where $l$ denotes the number of iterations of \ref{alg1:updateG}.

When we say that an algorithm is an NC algorithm. It should meet two requirements: first,
it works on a polynomial number of processors; second, it halts in poly-logarithmic time.
As discussed above, Algorithm~$\mathsf{A}_{bsc}$ meets the first requirement.
If we want to restrict Algorithm~$\mathsf{A}_{bsc}$ to be an NC algorithm,
we can make the number of iterations of \ref{alg1:updateG} to be poly-logarithmically bounded.
We use the symbol $\psi$ to denote a poly-logarithmically bounded function.
For any datalog program, if we restrict that the number of iterations of \ref{alg1:updateG} is bounded by $\psi(|\textbf{I}|)$,
the computing time of Algorithm~$\mathsf{A}_{bsc}$
is $c_1+\psi(|\textbf{I}|)\cdot c_2$. With this restriction, Algorithm~$\mathsf{A}_{bsc}$ is an NC algorithm denoted by
$\mathsf{A}_{bsc}^{\psi}$.

Based on $\mathsf{A}_{bsc}^{\psi}$, we can identify a class of datalog programs
$\mathcal{D}_{\mathsf{A}_{bsc}^{\psi}}$ such that all the datalog programs in it can be handled
by $\mathsf{A}_{bsc}^{\psi}$.
It is obvious that $\mathcal{D}_{\mathsf{A}_{bsc}^{\psi}}$ is a \texttt{PTD} class.
We use the following theorem to further show that this class can be captured based on
the materialization graphs of the datalog programs in $\mathcal{D}_{\mathsf{A}_{bsc}^{\psi}}$.

\begin{theorem}\label{theorem:a1}
For any datalog program $P$, $P\in\mathcal{D}_{\mathsf{A}_{bsc}^{\psi}}$ iff $P$ has a
materialization graph whose depth is upper-bounded by $\psi(|P|)$.
\end{theorem}

\begin{proof}[Proof sketch]
We can first prove that the number of iterations
of \ref{alg1:updateG} is actually the depth of the constructed materialization
graph. This theorem then follows by considering
Lemma~\ref{lemma:a1}.
\end{proof}

Consider Example~\ref{exp:mg} again. Let the integer $k$ be a variable. We can
get a class of datalog programs, denoted by $\mathbb{P}_{ex_1}$, where the rule set
is $\{R(x,y),A(y)\rightarrow A(x), S(x,y),R(y,z)\rightarrow R(x,z)\}$ and the fact
set varies according to $k$. The algorithm $\mathsf{A}_{bsc}^{\psi}$ is restricted in the sense that it cannot even work on the rather simple datalog program class $\mathbb{P}_{ex_1}$.
It can be checked that, for some materialization graph $\mathcal{G}_{ex_1}$
corresponding to a datalog program in $\mathbb{P}_{ex_1}$, \texttt{depth}($\mathcal{G}_{ex_1}$)$=k$ for some $k$.
This means that the depths of the materialization graphs are linearly bounded by $k$.
On the other hand, the sizes of the datalog programs in $\mathbb{P}_{ex_1}$ are polynomial of $k$.
Thus, for any $\psi$ that is poly-logarithmically bounded, we can always find a $k$ large
enough such that $\mathsf{A}_{bsc}^{\psi}$ terminates without constructing a materialization
graph for each datalog program in $\mathbb{P}_{ex_1}$.
However there indeed exists an NC algorithm that can handle $\mathbb{P}_{ex_1}$.
We discuss this in the next part.

\subsection{Optimizing Algorithm~$\mathsf{A}_{bsc}$ via the Single-Way Derivability}

In this part, we optimize Algorithm~$\mathsf{A}_{bsc}$ such that $P_{ex_1}$
can be handled. Based on the optimized variant of Algorithm~$\mathsf{A}_{bsc}$,
we can identify another \texttt{PTD} class.

We discuss our optimization based on a specific case in Example~\ref{exp:general}.
We find that, in this kind of case, the construction of a materialization graph can be accelerated.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.7\textwidth]{fig-general.eps}
\caption{A partial materialization graph.}
\label{fig:general}
\end{center}
\end{figure}

\begin{example}\label{exp:general}
Consider a snapshot of Algorithm~$\mathsf{A}_{bsc}$ in
Figure~\ref{fig:general}. A materialization graph $\mathcal{G}$
is being constructed for some datalog program $\langle R, \textbf{I}\rangle$.
The nodes in the dashed box denote the ones that have been added to $\mathcal{G}$.
In this snapshot, $v_0$ has been newly added
to $\mathcal{G}$ in the $l^{th}$ ($l\geq 1$) iteration.
Each of the nodes $v_i$ ($1\leq i\leq k$) has almost one parent node being not in $\mathcal{G}$,
while $v'$ has two parent nodes being not in $\mathcal{G}$.
All of the nodes $v_i$ ($1\leq i\leq k$) and $v'$ would be added to $\mathcal{G}$
afterwards.
\end{example}

In Example~\ref{exp:general}, $v_k$ would be added to $\mathcal{G}$ after \emph{at least} $k$
iterations by performing Algorithm~$\mathsf{A}_{bsc}$.
We can check that each edge $(v_{i-1},v_i)$ ($1\leq i\leq k$) in this stage follows the condition:
\emph{if the parent node $v_{i-1}$ is derivable, the child node $v_i$ has to be derivable} (this
is because that, for node $v_i$, node $v_{i-1}$ is the only parent node that has not been added to $\mathcal{G}$).
We call this condition \emph{the single-way derivability condition}, which intuitively says that the derivability of a
node only depends on one of its parent nodes.
Observe that node $v_k$ is reachable from node $v_0$ through the path $\tau=(v_0,v_1,...,v_k)$ where
each edge $(v_{i-1},v_i)$ ($1\leq i\leq k$) follows the single-way derivability condition;
we call such a path a \emph{single-way derivable} (SWD) path.
Further, the starting node $v_0$ in $\tau$ has been added to $\mathcal{G}$; in other words,
$v_0$ is derivable. Thus, all of the nodes $v_i$ ($1\leq i\leq k$) can be added to
$\mathcal{G}$ immediately.
Based on the above idea, we optimize Algorithm~$\mathsf{A}_{bsc}$ by checking whether such an SWD path
exits for some node $v$; if there is an SWD path for node $v$, this node can be added to $\mathcal{G}$
due to the single-way derivability condition.

Backing to Example~\ref{exp:general}, since
there exists an SWD path $(v_0,...,v_i)$ for each node $v_i$ ($1\leq i\leq k$), all of these nodes
can be added to $\mathcal{G}$ right after $v_0$. On the other hand,
node $v'$ has no SWD path, since it has two parent nodes being not in $\mathcal{G}$.

We next discuss how to determine the existence of SWD paths. Note that, SWD paths require us to
describe the reachability between two nodes. To this extent, we use a
binary transitive relation $\texttt{rch} \subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$,
e.g., \texttt{rch}$(v_1,v_2)$ means that $v_2$ is reachable from $v_1$.
In each iteration of \ref{alg1:updateG} in Algorithm~$\mathsf{A}_{bsc}$, we further compute a \texttt{rch}
relation (denoted by $S_{\textit{\tiny rch}}$) by performing the following process:

\begin{description}[leftmargin=2ex]
\item[(\textbf{\dag})] \emph{For each rule instantiation of the form $B_1,..,B_i,..,B_n\rightarrow H$
where $H$ has not been added to $\mathcal{G}$}:
\begin{enumerate}[leftmargin=2ex]
\item \emph{if the body atoms $B_1,...,B_n$ have all been added to $\mathcal{G}$, put \texttt{rch}$(B_1,H),...,$ \texttt{rch}$(B_n,H)$ in $S_{\textit{\tiny rch}}$};
\item \emph{if $B_i$ is the only node in the body that has not been added to $\mathcal{G}$,
    put \texttt{rch}$(B_i,H)$ in $S_{\textit{\tiny rch}}$}.\hfill$\Box$
\end{enumerate}
\end{description}

We then compute the transitive closure (denoted by $S^*_{\textit{\tiny rch}}$) with respect to
$S_{\textit{\tiny rch}}$. Based on the transitive closure, we can perform the optimization that,
for a node $v$, if there is a relation $\texttt{rch}(v',v)\in S^*_{\textit{\tiny rch}}$
where $v'$ has been added to $\mathcal{G}$, $v$ has an SWD path and can be added to $\mathcal{G}$.
The following algorithm applies this optimization strategy.\\

\noindent\texttt{Algorithm~$\mathsf{OPT}$}. The algorithm requires two inputs:
a datalog program $P=\langle R, \textbf{I}\rangle$ and a (partial) materialization graph $\mathcal{G}$ that is
being constructed from $P$. The following steps are performed:
\begin{enumerate}[leftmargin=6ex,label=(\textbf{\roman*})]
\item Compute a \texttt{rch} relation $S_{\textit{\tiny rch}}$ by following the above process (see $(\textbf{\dag})$).\label{rch}
\item Compute the transitive closure $S^*_{\textit{\tiny rch}}$ of $S_{\textit{\tiny rch}}$.\label{transClos}
\item Update $\mathcal{G}$ as follows: for any \texttt{rch}$(B_i,H)\in S_{\textit{\tiny rch}}$
that corresponds to `$B_1,..,B_i,..$ $,B_n\rightarrow H$'
    and there exists a node $B'$ such that \texttt{rch}$(B',B_i)\in S^*_{\textit{\tiny rch}}$ and $B'$ is
    in $\mathcal{G}$; if $H$ is not in $\mathcal{G}$ or $H$ is in $\mathcal{G}$ but has no parent pointing
    to it, add $H$ and $B_i$ (if $B_i$ is not in $\mathcal{G}$) to $\mathcal{G}$, and create the edges
    $e(B_1, H),...,e(B_n, H)$ in $\mathcal{G}$. Do nothing for other statements
    \texttt{rch}$(B_j,H)\in S_{\textit{\tiny rch}}$.\label{updateG}\hfill$\Box$
\end{enumerate}

It is well known that there is an NC algorithm for computing the
transitive closure \cite{Allender07}.
Based on this result and Algorithm~$\mathsf{OPT}$, we propose an optimized variant of Algorithm~$\mathsf{A}_{bsc}$:\\

\noindent\texttt{Algorithm~$\mathsf{A}_{opt}$}. Given a datalog program $P=\langle R, \textbf{I}\rangle$, the algorithm
returns a materialization graph $\mathcal{G}$ of $P$.
Suppose we have $|P^*|$ processors, and each rule instantiation in $P^*$ is
assigned to one processor.
Initially $\mathcal{G}$ is empty. The following steps are then performed:
\begin{enumerate}[leftmargin=8ex,label=(\textit{Step \arabic*}),ref=Step~\arabic*]
\item Add all facts in $\textbf{I}$ to $\mathcal{G}$.\label{alg3:addFacts}
\item Compute $S_{\textit{\tiny rch}}$ by performing \ref{rch} in Algorithm~$\mathsf{OPT}$; use an NC
    algorithm to compute the transitive closure $S^*_{\textit{\tiny rch}}$ (see \ref{transClos} in Algorithm~$\mathsf{OPT}$);
    update $\mathcal{G}$ by performing \ref{updateG}  in Algorithm~$\mathsf{OPT}$.\label{alg3:updateG}
\item If no node has been added to $\mathcal{G}$ (in \ref{alg3:updateG}), terminate,
    otherwise iterate \ref{alg3:updateG}. \label{alg3:halt}\hfill$\Box$
\end{enumerate}

It should be noted that there has to be an SWD path for any derivable node in some iteration when performing
Algorithm~$\mathsf{A}_{opt}$.
The following lemma shows the correctness of Algorithm~$\mathsf{A}_{opt}$.

\begin{lemma}\label{lemma:a3}
Given a datalog program $P=\langle R, \textbf{I}\rangle$,
$\mathsf{A}_{opt}$ halts and outputs a materialization graph $\mathcal{G}$ of $P$.
\end{lemma}

\begin{example}\label{exp:opt}
We perform Algorithm~$\mathsf{A}_{opt}$ on the datalog program $P_{ex_1}$ in Example~\ref{exp:mg}.
Initially, $R(a_1,b)$ is in the materialization graph $\mathcal{G}_{ex_1}$.
In the first iteration of \ref{alg3:updateG}, all the rule instantiations are in two kinds of forms:
`$R(a_i,b),A(b)\rightarrow A(a_i)$' and `$S(a_i,a_{i-1}),$ $R(a_{i-1},b)\rightarrow R(a_i,b)$' $(2\leq i\leq k)$,
$S_{\textit{\tiny rch}}$ is the set $\{$\texttt{rch}$(R(a_{i-1},b), R(a_i,b))|2\leq i\leq k\}
\cup\{$\texttt{rch}$(R(a_i,b), A(a_i))|1\leq i\leq k\}$.
In the transitive closure of $S_{\textit{\tiny rch}}$,
one can check that \texttt{rch}$(R(a_1,b), R(a_i,b)),$
\texttt{rch}$(R(a_1,b), A(a_i))\in S^*_{\textit{\tiny rch}}(2\leq i\leq k)$.
Thus, $R(a_i,b)$ and $A(a_i)$ ($2\leq i\leq k$) can all be added to $\mathcal{G}_{ex_1}$
in the first iteration of \ref{alg3:updateG}.
\end{example}

>>>>

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "parallel-tractability-J"
%%% End:
